A pipeline to analyse time-course gene expression data

### 2020-07-21 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@

The manuscript was last compile on 2020-07-04 10:52:02.

@@

\texttt{moanin} is available from GitHub, and can be installed using the \texttt{install\_github} function in the package \texttt{devtools}:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(devtools)}
\KeywordTok{install_github}\NormalTok{(}\StringTok{"NelleV/moanin"}\NormalTok{)}
\KeywordTok{library}\NormalTok{(moanin)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# From github}
\KeywordTok{library}\NormalTok{(timecoursedata)}

@@

\CommentTok{# From Bioconductor}
\KeywordTok{library}\NormalTok{(topGO)}
\KeywordTok{library}\NormalTok{(biomaRt)}
\KeywordTok{library}\NormalTok{(KEGGprofile)}
\KeywordTok{library}\NormalTok{(BiocWorkflowTools)}

\CommentTok{# From GitHub}
\KeywordTok{library}\NormalTok{(moanin)}
\end{Highlighting}
\end{Shaded}

\textbf{EAP}: Why is this list not the same as the (hidden via echo=FALSE) list at
the beginning of the document? I also needed to install \texttt{BiocWorkflowTools},
\texttt{KEGGprofile} to run the .Rmd
\textbf{NV}: I think some of the libraries above are not needed anymore. Some are
also only used in hidden elements. I just checked that the whole pipeline
works, by executing all command lines the reader will see, and the libraries
are all loaded fine. I think BiocWorkflowTools is only need in the compilation
of the Rmd file.

@@

In both
situations, we first need to set up a object (a \texttt{moanin\_model} object) to
hold the meta data, as well as information for fitting the spline model
(formula, number of degrees of freedom of the splines, \ldots{}). The \texttt{moanin\_model} object
will contain information used
throughout this analysis, in particular the condition and timepoints of each sample and the
the basis matrix for fitting splines models.

We start by creating the \texttt{moanin\_model} object using the \texttt{create\_moanin\_model}
function. We need to provide two things to the function: a \texttt{data.frame} with
the metadata, and the number of degrees of freedom of the splines to be used in the
functional modeling. The metadata \texttt{data.frame} object should contain at least
two columns: one named \texttt{Group}, containing the treatement effect, and a
second one named \texttt{Timepoint} containing the timepoint information.

@@

We create a \texttt{moanin\_model} for our data:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{moanin_model =}\StringTok{ }\KeywordTok{create_moanin_model}\NormalTok{(meta, }\DataTypeTok{degrees_of_freedom=}\DecValTok{6}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{print}\NormalTok{(moanin_model)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## moanin_model object on 209 samples containing the following information:
## 1) Meta data with 5 variables:
## [1] ""            "Group"       "Replicate"   "Timepoint"   "WeeklyGroup"
## 2) Basis matrix with 35 basis functions
## Basis matrix was constructed with the following formula
## ~Group + Group:splines::ns(Timepoint, df = degrees_of_freedom) + 0 
## Where degrees_of_freedom=6 
## To access these objects use:
##  <x_name>$degrees_of_freedom
##  <x_name>$basis
##  <x_name>$meta
##  <x_name>$formula
\end{verbatim}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Define contrasts  }
\NormalTok{contrasts =}\StringTok{ }\KeywordTok{create_timepoints_contrasts}\NormalTok{(}\StringTok{"M"}\NormalTok{, }\StringTok{"VL"}\NormalTok{, moanin_model)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{weekly_de_analysis =}\StringTok{ }\KeywordTok{DE_timepoints}\NormalTok{(}
\NormalTok{     data, moanin_model, contrasts,}
     \DataTypeTok{use_voom_weights=}\OtherTok{FALSE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\CommentTok{# The function takes the data (data.frame or named matrix), the meta data}
\CommentTok{# (data.frame containing a timepoint and group column, the first corresponding}
\CommentTok{# to the time-course information, the latter corresponding to the}
\CommentTok{# treatment).}
\NormalTok{DE_results =}\StringTok{ }\KeywordTok{DE_timecourse}\NormalTok{(}
\NormalTok{    data, moanin_model, timecourse_contrasts,}
    \DataTypeTok{use_voom_weights=}\OtherTok{FALSE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

Computing the log fold change on a time-course experiment is not trivial: one
can be interested in the average log-fold change across time, or the
cumulative log-fold change. Sometimes a gene can be over-expressed at the
beginning of the time-course data, and then under-expressed at the end of the
experiment. As a result, \texttt{moanin} provides a number of possible ways to
compute the log fold change across the whole time-course. This is done via the function \texttt{estimate\_log\_fold\_change} which takes as
arguments the data, the \texttt{moanin\_model} object, the contrasts to evaluate, and the method to use to estimate the log-fold change.

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{log_fold_change_timepoints =}\StringTok{ }\KeywordTok{estimate_log_fold_change}\NormalTok{(}
\NormalTok{    data, moanin_model, timecourse_contrasts,  }\DataTypeTok{method=}\StringTok{"timely"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{tabular}{lrrrrr}
\toprule
  & M-K:1 & M-K:7 & M-K:11 & M-K:14 & M-K:2\\
\midrule
NM\_009912 & -0.0061424 & 0.0224948 & -0.0515947 & -0.1143973 & 0.0614619\\
NM\_008725 & 2.6547855 & 0.2734387 & -0.6882925 & 0.4718807 & -1.7463319\\
NM\_007473 & -0.1603623 & 0.1017655 & 0.5079573 & 0.0595773 & 0.4250185\\
ENSMUST00000094955 & 0.3505920 & 0.6463011 & 0.2564796 & 0.3599102 & 0.4177116\\
NM\_001042489 & 0.0889750 & 0.2321381 & -0.1891501 & 0.1786113 & 0.3320644\\
\addlinespace
NM\_008159 & -0.3863770 & 0.0298395 & -0.0862755 & -0.3266351 & -0.1893062\\
\bottomrule
\end{tabular}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{log_fold_change_timecourse =}\StringTok{ }\KeywordTok{estimate_log_fold_change}\NormalTok{(}
\NormalTok{    data, moanin_model, timecourse_contrasts,  }\DataTypeTok{method=}\StringTok{"timecourse"}\NormalTok{)}

\NormalTok{log_fold_change_sum =}\StringTok{ }\KeywordTok{estimate_log_fold_change}\NormalTok{(}
\NormalTok{    data, moanin_model, timecourse_contrasts,  }\DataTypeTok{method=}\StringTok{"sum"}\NormalTok{)}

\NormalTok{log_fold_change_max =}\StringTok{ }\KeywordTok{estimate_log_fold_change}\NormalTok{(}
\NormalTok{    data, moanin_model, timecourse_contrasts, }\DataTypeTok{method=}\StringTok{"max"}\NormalTok{)}

\NormalTok{log_fold_change_min =}\StringTok{ }\KeywordTok{estimate_log_fold_change}\NormalTok{(}
\NormalTok{    data, moanin_model, timecourse_contrasts, }\DataTypeTok{method=}\StringTok{"min"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{tabular}{lrrrr}
\toprule
  & M-K & M-C & M-VL & M-VH\\
\midrule
NM\_009912 & 0.432 & 0.753 & -0.836 & -0.488\\
NM\_008725 & -1.257 & -2.008 & 1.924 & 1.729\\
NM\_007473 & 0.385 & 0.335 & 0.516 & 0.263\\
ENSMUST00000094955 & 0.355 & 0.376 & -0.316 & 0.256\\
NM\_001042489 & -0.266 & 0.410 & 0.415 & 0.374\\
\bottomrule
\end{tabular}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{top_DE_genes_pval =}\StringTok{ }\KeywordTok{names}\NormalTok{(}\KeywordTok{sort}\NormalTok{(pvalue)[}\DecValTok{1}\OperatorTok{:}\DecValTok{10}\NormalTok{])}
\KeywordTok{plot_splines_data}\NormalTok{(data[top_DE_genes_pval, ], moanin_model,}
    \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group, }\DataTypeTok{smooth=}\OtherTok{TRUE}\NormalTok{,}
    \DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{1.5}\NormalTok{,}\FloatTok{2.5}\NormalTok{,}\DecValTok{2}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{top_DE_genes_lfc =}\StringTok{ }\KeywordTok{names}\NormalTok{(}
    \KeywordTok{sort}\NormalTok{(}\KeywordTok{abs}\NormalTok{(lfc_timecourse),}
     \DataTypeTok{decreasing=}\OtherTok{TRUE}\NormalTok{)[}\DecValTok{1}\OperatorTok{:}\DecValTok{10}\NormalTok{])}
\KeywordTok{plot_splines_data}\NormalTok{(data[top_DE_genes_lfc, ], moanin_model,}
    \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group, }\DataTypeTok{smooth=}\OtherTok{TRUE}\NormalTok{,}
    \DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{1.5}\NormalTok{,}\FloatTok{2.5}\NormalTok{,}\DecValTok{2}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

\NormalTok{genes_to_keep =}\StringTok{ }\KeywordTok{row.names}\NormalTok{(}
\NormalTok{    log_fold_change_max[}
\NormalTok{    (}\KeywordTok{rowSums}\NormalTok{(log_fold_change_max }\OperatorTok{>}\StringTok{ }\DecValTok{2}\NormalTok{) }\OperatorTok{>}\StringTok{ }\DecValTok{0}\NormalTok{) }\OperatorTok{&}
\StringTok{    }\NormalTok{(fishers_qval }\OperatorTok{<}\StringTok{ }\FloatTok{0.05}\NormalTok{), ])}
\CommentTok{# Keep the data corresponding to the genes of interest in another variable.}
\NormalTok{y =}\StringTok{ }\KeywordTok{as.matrix}\NormalTok{(data[genes_to_keep, ])}
\end{Highlighting}
\end{Shaded}

@@

After filtering, we are left with 3950 genes. We can then apply a
clustering routine. As observed by looking at genes found differentially expressed,
many genes share a similar gene expression pattern, but on different scale.

@@

\begin{itemize}
\tightlist
\item
  \textbf{Splines estimation}: for each gene, fit the splines function with the basis
  of your choice (as contained in the \texttt{moanin\_model} object).
\item
  \textbf{Rescaling splines}: for each gene, rescale the estimated splines function
  such that the values are bounded between 0 and 1 and thus comparable between genes.
\item
  \textbf{K-means}: apply k-means on the rescaled fitted values of the splines to estimate the
  cluster centroids.
\end{itemize}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# First fit the kmeans clusters}
\NormalTok{kmeans_clusters =}\StringTok{ }\KeywordTok{splines_kmeans}\NormalTok{(}
\NormalTok{    y, moanin_model, }\DataTypeTok{n_clusters=}\DecValTok{8}\NormalTok{,}
    \DataTypeTok{random_seed=}\DecValTok{42}\NormalTok{,}
    \DataTypeTok{n_init=}\DecValTok{20}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{plot_splines_data}\NormalTok{(}
\NormalTok{    kmeans_clusters}\OperatorTok{$}\NormalTok{centroids, moanin_model,}
    \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group,}
    \DataTypeTok{smooth=}\OtherTok{TRUE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{cluster_of_interest =}\StringTok{ }\DecValTok{2}
\NormalTok{cluster2Genes =}\StringTok{ }\KeywordTok{names}\NormalTok{(}
\NormalTok{    kmeans_clusters}\OperatorTok{$}\NormalTok{clusters[kmeans_clusters}\OperatorTok{$}\NormalTok{clusters}\OperatorTok{==}\NormalTok{cluster_of_interest])}
\NormalTok{dataToPlot=}\KeywordTok{rbind}\NormalTok{(y[cluster2Genes[}\DecValTok{3}\OperatorTok{:}\DecValTok{6}\NormalTok{], ],}\StringTok{"Centroid"}\NormalTok{=kmeans_clusters}\OperatorTok{$}\NormalTok{centroids[cluster_of_interest,])}
\KeywordTok{plot_splines_data}\NormalTok{(dataToPlot, moanin_model,}
    \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group, }\DataTypeTok{smooth=}\OtherTok{TRUE}\NormalTok{,}\DataTypeTok{simpleY =}\OtherTok{FALSE}\NormalTok{,}
    \DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{1.5}\NormalTok{,}\FloatTok{2.5}\NormalTok{,}\DecValTok{2}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/genesVsCentroids-1} \end{center}

@@

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Then assign scores and labels to all the data, using a goodness-of-fit}
\CommentTok{# scoring function.}
\NormalTok{scores_and_labels =}\StringTok{ }\KeywordTok{splines_kmeans_score_and_label}\NormalTok{(}
\NormalTok{    data, kmeans_clusters)}
\end{Highlighting}
\end{Shaded}

The \texttt{scores\_and\_labels} list contains two elements:

\begin{verbatim}
- `scores`: the matrix of shape `n_cluster` x `n_genes`,
  containing for each gene and each cluster, the goodness of fit as
  described above.
- `labels`: the labels for all of the genes with a
  sufficiently good goodness-of-fit score.
\end{verbatim}

>>>>

%% added by MLS, this is not in the F1000 style by default %%

\hypersetup{unicode=true,
            pdftitle={A pipeline to analyse time-course gene expression data},
            pdfkeywords={time-course gene expression data, clustering, differential expression, workflow},
            pdfborder={0 0 0},
            breaklinks=true}

%% End added by MLS %%

@@

The manuscript was last compile on 2020-07-21 12:22:19.

@@

\texttt{moanin} is available from GitHub, and can be installed using the \texttt{install\_github} function in the package \texttt{devtools}, along with the corresponding package that contains time course datasets we will use:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(devtools)}
\KeywordTok{install_github}\NormalTok{(}\StringTok{"NelleV/moanin"}\NormalTok{)}
\KeywordTok{install_github}\NormalTok{(}\StringTok{"NelleV/timecoursedata"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# From Github}
\KeywordTok{library}\NormalTok{(moanin)}
\KeywordTok{library}\NormalTok{(timecoursedata)}

@@

\CommentTok{# From Bioconductor}
\KeywordTok{library}\NormalTok{(topGO)}
\KeywordTok{library}\NormalTok{(biomaRt)}
\KeywordTok{library}\NormalTok{(KEGGprofile)}
\KeywordTok{library}\NormalTok{(BiocWorkflowTools) }\CommentTok{#Needed for compiling the .Rmd script}
\end{Highlighting}
\end{Shaded}

If Bioconductor is installed, the CRAN and Bioconductor packages above can be installed via

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{BiocManager}\OperatorTok{::}\KeywordTok{install}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\StringTok{"NMF"}\NormalTok{,}\StringTok{"ggfortify"}\NormalTok{,}\StringTok{"topGO"}\NormalTok{,}\StringTok{"biomaRt"}\NormalTok{,}\StringTok{"KEGGprofile"}\NormalTok{,}\StringTok{"BiocWorkflowTools"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

In both
situations, we first need to set up a object (a \texttt{Moanin} object) to
hold the meta data, as well as information for fitting the spline model
(formula, number of degrees of freedom of the splines, \ldots{}). The \texttt{Moanin} object
will contain information used
throughout this analysis, in particular the condition and timepoints of each sample and the
the basis matrix for fitting splines models.

We start by creating the \texttt{Moanin} object using the \texttt{create\_moanin\_model}
function. We need to provide two things to the function: a \texttt{data.frame} with
the metadata, and the number of degrees of freedom of the splines to be used in the
functional modeling. The metadata \texttt{data.frame} object should contain at least
two columns: one named \texttt{Group}, containing the treatement effect, and a
second one named \texttt{Timepoint} containing the timepoint information.

@@

We create a \texttt{Moanin} object for our data:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{moanin_model =}\StringTok{ }\KeywordTok{create_moanin_model}\NormalTok{(}\DataTypeTok{data=}\NormalTok{data, }\DataTypeTok{meta=}\NormalTok{meta,}
    \DataTypeTok{degrees_of_freedom=}\DecValTok{6}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{show}\NormalTok{(moanin_model)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Moanin object on 209 samples containing the following information:
## Group variable given by 'Group' with the following levels:
##  M  K  C VL VH 
## 42 42 42 42 41 
## Time variable given by 'Timepoint'
## Basis matrix with 35 basis_matrix functions
## Basis matrix was constructed with the following spline_formula
## ~Group + Group:splines::ns(Timepoint, df = 6) + 0 
## 
## Information about the data (a SummarizedExperiment object):
## class: SummarizedExperiment 
## dim: 39544 209 
## metadata(0):
## assays(1): ''
## rownames(39544): NM_009912 NM_008725 ... NM_010201.1 AK078781
## rowData names(0):
## colnames(209): GSM1557140 GSM1557141 ... GSM1557347 GSM1557348
## colData names(5): '' Group Replicate Timepoint WeeklyGroup
\end{verbatim}

The \texttt{Moanin} class extends the \texttt{SummarizedExperiment} class of Bioconductor, so that the data as well as the meta data and our spline information are all held in one object. This means that the object can be subsetted just like the data matrix, and the corresponding meta data and basis function evaluations will be similarly subsetted:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{dim}\NormalTok{(moanin_model)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 39544   209
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{moanin_model[}\DecValTok{1}\OperatorTok{:}\DecValTok{10}\NormalTok{,}\DecValTok{1}\OperatorTok{:}\DecValTok{3}\NormalTok{]}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Moanin object on 3 samples containing the following information:
## Group variable given by 'Group' with the following levels:
##  M  K  C VL VH 
##  0  3  0  0  0 
## Time variable given by 'Timepoint'
## Basis matrix with 35 basis_matrix functions
## Basis matrix was constructed with the following spline_formula
## ~Group + Group:splines::ns(Timepoint, df = 6) + 0 
## 
## Information about the data (a SummarizedExperiment object):
## class: SummarizedExperiment 
## dim: 10 3 
## metadata(0):
## assays(1): ''
## rownames(10): NM_009912 NM_008725 ... NM_013782 NM_028622
## rowData names(0):
## colnames(3): GSM1557140 GSM1557141 GSM1557142
## colData names(5): '' Group Replicate Timepoint WeeklyGroup
\end{verbatim}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Define contrasts  }
\NormalTok{contrasts =}\StringTok{ }\KeywordTok{create_timepoints_contrasts}\NormalTok{(moanin_model,}\StringTok{"M"}\NormalTok{, }\StringTok{"VL"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{weekly_de_analysis =}\StringTok{ }\KeywordTok{DE_timepoints}\NormalTok{( moanin_model, contrasts,}
     \DataTypeTok{use_voom_weights=}\OtherTok{FALSE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\CommentTok{# The function takes the data (data.frame or named matrix), the meta data}
\CommentTok{# (data.frame containing a timepoint and group column, the first corresponding}
\CommentTok{# to the time-course information, the latter corresponding to the}
\CommentTok{# treatment).}
\NormalTok{DE_results =}\StringTok{ }\KeywordTok{DE_timecourse}\NormalTok{( moanin_model, timecourse_contrasts,}
    \DataTypeTok{use_voom_weights=}\OtherTok{FALSE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

Computing the log fold change on a time-course experiment is not trivial: one
can be interested in the average log-fold change across time, or the
cumulative log-fold change. Sometimes a gene can be over-expressed at the
beginning of the time-course data, and then under-expressed at the end of the
experiment. As a result, \texttt{moanin} provides a number of possible ways to
compute the log fold change across the whole time-course. This is done via the function \texttt{estimate\_log\_fold\_change} which takes as
arguments the data, the \texttt{Moanin} object, the contrasts to evaluate, and the method to use to estimate the log-fold change.

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{log_fold_change_timepoints =}\StringTok{ }\KeywordTok{estimate_log_fold_change}\NormalTok{(}
\NormalTok{    moanin_model, timecourse_contrasts,  }\DataTypeTok{method=}\StringTok{"timely"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{tabular}{lrrrrr}
\toprule
  & M-K:0 & M-K:3 & M-K:6 & M-K:9 & M-K:12\\
\midrule
NM\_009912 & -0.5662398 & -0.2454676 & -0.4096932 & -0.3320085 & -0.0502552\\
NM\_008725 & 0.9947059 & -3.4609546 & 0.6561484 & -1.2106329 & -0.1998796\\
NM\_007473 & -0.5176866 & -0.3146529 & 0.2327960 & -0.3445581 & 0.0476160\\
ENSMUST00000094955 & -0.2201803 & 0.2566677 & -0.0358376 & -0.0469726 & -0.0707469\\
NM\_001042489 & -0.2475888 & -0.2043850 & -0.3428368 & -0.3930846 & -0.4821660\\
\addlinespace
NM\_008159 & -0.7466435 & -0.0678902 & -0.3072220 & -0.7067393 & -0.4966867\\
\bottomrule
\end{tabular}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{log_fold_change_timecourse =}\StringTok{ }\KeywordTok{estimate_log_fold_change}\NormalTok{(}
\NormalTok{    moanin_model, timecourse_contrasts,  }\DataTypeTok{method=}\StringTok{"timecourse"}\NormalTok{)}

\NormalTok{log_fold_change_sum =}\StringTok{ }\KeywordTok{estimate_log_fold_change}\NormalTok{(}
\NormalTok{    moanin_model, timecourse_contrasts,  }\DataTypeTok{method=}\StringTok{"sum"}\NormalTok{)}

\NormalTok{log_fold_change_max =}\StringTok{ }\KeywordTok{estimate_log_fold_change}\NormalTok{(}
\NormalTok{    moanin_model, timecourse_contrasts, }\DataTypeTok{method=}\StringTok{"max"}\NormalTok{)}

\NormalTok{log_fold_change_min =}\StringTok{ }\KeywordTok{estimate_log_fold_change}\NormalTok{(}
\NormalTok{    moanin_model, timecourse_contrasts, }\DataTypeTok{method=}\StringTok{"min"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{tabular}{lrrrr}
\toprule
  & M-K & M-C & M-VL & M-VH\\
\midrule
NM\_009912 & -0.409 & -0.753 & -1.025 & -1.420\\
NM\_008725 & 1.575 & 2.008 & 2.543 & 2.618\\
NM\_007473 & -0.416 & -0.335 & -0.213 & 0.364\\
ENSMUST00000094955 & -0.113 & -0.376 & -0.156 & -0.309\\
NM\_001042489 & -0.234 & -0.410 & 0.308 & 0.430\\
\bottomrule
\end{tabular}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{top_DE_genes_pval =}\StringTok{ }\KeywordTok{names}\NormalTok{(}\KeywordTok{sort}\NormalTok{(pvalue)[}\DecValTok{1}\OperatorTok{:}\DecValTok{10}\NormalTok{])}
\KeywordTok{plot_splines_data}\NormalTok{(moanin_model,}\DataTypeTok{subset_data=}\NormalTok{top_DE_genes_pval, }
    \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group, }\DataTypeTok{smooth=}\OtherTok{TRUE}\NormalTok{,}
    \DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{1.5}\NormalTok{,}\FloatTok{2.5}\NormalTok{,}\DecValTok{2}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{top_DE_genes_lfc =}\StringTok{ }\KeywordTok{names}\NormalTok{(}
    \KeywordTok{sort}\NormalTok{(}\KeywordTok{abs}\NormalTok{(lfc_timecourse),}
     \DataTypeTok{decreasing=}\OtherTok{TRUE}\NormalTok{)[}\DecValTok{1}\OperatorTok{:}\DecValTok{10}\NormalTok{])}
\KeywordTok{plot_splines_data}\NormalTok{(moanin_model, }\DataTypeTok{subset_data=}\NormalTok{top_DE_genes_lfc, }
    \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group, }\DataTypeTok{smooth=}\OtherTok{TRUE}\NormalTok{,}
    \DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{1.5}\NormalTok{,}\FloatTok{2.5}\NormalTok{,}\DecValTok{2}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

\NormalTok{genes_to_keep =}\StringTok{ }\KeywordTok{row.names}\NormalTok{(}
\NormalTok{    log_fold_change_max[}
\NormalTok{    (}\KeywordTok{rowSums}\NormalTok{(log_fold_change_max }\OperatorTok{>}\StringTok{ }\DecValTok{2}\NormalTok{) }\OperatorTok{>}\StringTok{ }\DecValTok{0}\NormalTok{) }\OperatorTok{&}
\StringTok{    }\NormalTok{(fishers_qval }\OperatorTok{<}\StringTok{ }\FloatTok{0.05}\NormalTok{), ])}
\CommentTok{# Keep the data corresponding to the genes of interest in another variable.}
\CommentTok{# by subsetting the `moanin_model`, which contains the data.}
\NormalTok{de_moanin_model =}\StringTok{ }\NormalTok{moanin_model[genes_to_keep,]}
\end{Highlighting}
\end{Shaded}

@@

After filtering, we are left with 5521 genes. We can then apply a
clustering routine. As observed by looking at genes found differentially expressed,
many genes share a similar gene expression pattern, but on different scale.

@@

\begin{itemize}
\tightlist
\item
  \textbf{Splines estimation}: for each gene, fit the splines function with the basis
  of your choice (as contained in the \texttt{Moanin} object).
\item
  \textbf{Rescaling splines}: for each gene, rescale the estimated splines function
  such that the values are bounded between 0 and 1 and thus comparable between genes.
\item
  \textbf{K-means}: apply k-means on the rescaled fitted values of the splines to estimate the
  cluster centroids.
\end{itemize}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# First fit the kmeans clusters}
\NormalTok{kmeans_clusters =}\StringTok{ }\KeywordTok{splines_kmeans}\NormalTok{(de_moanin_model, }\DataTypeTok{n_clusters=}\DecValTok{8}\NormalTok{,}
    \DataTypeTok{random_seed=}\DecValTok{42}\NormalTok{,}
    \DataTypeTok{n_init=}\DecValTok{20}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{plot_splines_data}\NormalTok{(de_moanin_model,}
    \DataTypeTok{data=}\NormalTok{kmeans_clusters}\OperatorTok{$}\NormalTok{centroids, }
    \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group,}
    \DataTypeTok{smooth=}\OtherTok{TRUE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{cluster_of_interest =}\StringTok{ }\DecValTok{2}
\NormalTok{cluster2Genes =}\StringTok{ }\KeywordTok{names}\NormalTok{(}
\NormalTok{    kmeans_clusters}\OperatorTok{$}\NormalTok{clusters[kmeans_clusters}\OperatorTok{$}\NormalTok{clusters}\OperatorTok{==}\NormalTok{cluster_of_interest])}
\CommentTok{# }\AlertTok{FIXME}\CommentTok{ This line fails}
\CommentTok{# dataToPlot = rbind(}
\CommentTok{#     assay(de_moanin_model[cluster2Genes[3:6],]), }
\CommentTok{#    "Centroid"=kmeans_clusters$centroids[cluster_of_interest,])}
\CommentTok{# plot_splines_data(de_moanin_model, data=dataToPlot, }
\CommentTok{#     colors=ann_colors$Group, smooth=TRUE, simpleY=FALSE,}
\CommentTok{#    mar=c(1.5,2.5,2,0.1))}
\end{Highlighting}
\end{Shaded}

@@

We can also overlay the centroid on the data, after rescaling the data to be between 0-1, like the centroids

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{de_moanin_rescale =}\StringTok{ }\NormalTok{de_moanin_model}
\KeywordTok{assay}\NormalTok{(de_moanin_rescale) =}\StringTok{ }\KeywordTok{rescale_values}\NormalTok{(}\DataTypeTok{data=}\KeywordTok{assay}\NormalTok{(de_moanin_model))}
\KeywordTok{plot_splines_data}\NormalTok{(de_moanin_rescale,  }
    \DataTypeTok{centroid=}\NormalTok{kmeans_clusters}\OperatorTok{$}\NormalTok{centroids[cluster_of_interest,], }
    \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group, }\DataTypeTok{smooth=}\OtherTok{TRUE}\NormalTok{,}\DataTypeTok{simpleY =}\OtherTok{FALSE}\NormalTok{,}
    \DataTypeTok{subset_data=}\NormalTok{cluster2Genes[}\DecValTok{3}\OperatorTok{:}\DecValTok{6}\NormalTok{],}
    \DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{1.5}\NormalTok{,}\FloatTok{2.5}\NormalTok{,}\DecValTok{2}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/unnamed-chunk-19-1} \end{center}

\textbf{EAP} I added the functionality to plot centroid over data, rather than just fit to data, but I'm not sure it makes much sense to include this. The \texttt{rescale\_values} function is too sensitive to outliers. Would be better to rescale each centroid to scale of gene. If we had a helper function to do that, could easily add that to plot\_splines\_data (made issue on moanin), but probably right now should remove this.

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Then assign scores and labels to *all* the data, using a goodness-of-fit}
\CommentTok{# scoring function.}
\NormalTok{scores_and_labels =}\StringTok{ }\KeywordTok{splines_kmeans_score_and_label}\NormalTok{(}
\NormalTok{    moanin_model, kmeans_clusters)}
\end{Highlighting}
\end{Shaded}

The \texttt{scores\_and\_labels} list contains three elements:

\begin{verbatim}
- `scores`: the matrix of shape `n_cluster` x `n_genes`,
  containing for each gene and each cluster, the goodness of fit as
  described above.
- `labels`: the labels for all of the genes with a
  sufficiently good goodness-of-fit score.
- `score_cutoff`: the cutoff used on `scores` to determine whether to 
  assign a label
\end{verbatim}

\textbf{Assigning cluster labels:} We could just assign each gene to a cluster based on which cluster gave the minimum score. By default, \texttt{splines\_kmeans\_score\_and\_label} does not do that, but rather requires a sufficiently low enough goodness-of-fit score. The criteria for being ``sufficiently low'' is based on looking at the distribution of the scores of all genes on all clusters (i.e.~the entire \texttt{scores} matrix returned by \texttt{splines\_kmeans\_score\_and\_label}). A gene is then assigned to a cluster only if their best score is above the 50\% percentile of that distribution, with the remaining genes getting \texttt{NA} as their assignment (this choice can be changed by \texttt{proportion\_genes\_to\_label}). Note that because the distribution of scores of genes across \emph{all} clusters are used, this is not equivalent to assigning 50\% of genes to a cluster -- it is possible that all genes are assigned to a cluster.
